---
title: "Study of delay prediction in the US airport network"
author: "Kerim Kiliç"
subtitle: "Supervised Machined Learning using flight data"
output:
  html_document:
    df_print: paged
    toc: true
    toc_depth: 2
    number_sections: true
    toc_float: true
---

# Libraries

The following three libraries are used in this R markdown file.

```{r setup, message=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(data.table)
library(tidymodels)
library(sparklyr)
```

# Initialize spark and read in data

```{r}
### Uncomment these two lines to install spark, 
### and optionally to disconnect an existing connection
# spark_install(version = "3.3.0")
# spark_disconnect(sc)
sc <- spark_connect(master = "local")

### Uncomment these two lines to read in the RDS file of the data of 2017
# my_data <- readRDS("flights_2017.RDS")
# fwrite(my_data,file = "flights_2017.csv")
raw_data <- spark_read_csv(sc,"flights_2017.csv")
raw_data %>% glimpse()
```

# Clean and classify data

```{r}
### Remove redundant variables and classify the data based on the numerical delays
spark_data <- raw_data %>%
  select(-id,-year,-wheels_on_time,-wheels_off_time,-planned_departure_time,-planned_arrival_time,
         -actual_departure_time,-actual_arrival_time,-tail_number,-flight_number,-carrier) %>%
  mutate(delay = case_when(carrier_delay + weather_delay + nas_delay + security_delay + late_aircraft_delay > 0 ~ "1",
                           carrier_delay + weather_delay + nas_delay + security_delay + late_aircraft_delay == 0 ~ "0"))

### Remove the numerical delays since we only look at classification for now
spark_data <- spark_data %>%
  select(-carrier_delay,-weather_delay,-nas_delay,-security_delay,-late_aircraft_delay)
### Remove rows that contain any number of NA values
spark_data <- na.omit(spark_data)
spark_data %>% glimpse()

### Index the categorical features and apply one hot encoding
spark_data <- ft_string_indexer(spark_data, input_col = "origin", output_col = "origin_ind")
spark_data <- ft_string_indexer(spark_data, input_col = "destination", output_col = "destination_ind")
spark_data_one_hot_encoded <- ft_one_hot_encoder(spark_data, input_cols = c("origin_ind","destination_ind"), output_cols = c("origin_out","destination_out")) %>%
  select(-origin,-destination)
spark_data_one_hot_encoded %>% glimpse()
```

# Splitting data in train and test sets

Take a sample set of the data to reduce time to train the model since we are just playing around with the package.

```{r}
train_data <- spark_data_one_hot_encoded %>%
  group_by(delay) %>%
  sample_frac(0.0008)

test_data <- spark_data_one_hot_encoded %>%
  group_by(delay) %>%
  sample_frac(0.0002)

train_data %>% glimpse()
```

# Train a logistic regression model

Train a logistic regression model using the train dataset and get model performance on training data.

```{r}
### Train a logistic regression model ###
glm_model <- ml_logistic_regression(train_data, "delay ~ .")
### Performance on train set
augment_value <- augment(glm_model) %>% collect()
augment_value$delay <- as.factor(augment_value$delay)
augment_value$.predicted_label <- as.factor(augment_value$.predicted_label)
augment_value %>%
  accuracy(estimate = .predicted_label, truth = delay)
```

Make a prediction based on test data and get model performance.

```{r}
### Performance on test set
ml_predict(glm_model,test_data) %>%
  ml_metrics_multiclass()
```

# Train a decision tree model

Train a decision tree model using the train dataset and get model performance on training data.

```{r}
### Train a decision tree model ###
tree_model <- ml_decision_tree_classifier(train_data, "delay ~ .", max_bins = 317)
### Performance on train set
augment_value <- augment(tree_model) %>% collect()
augment_value$delay <- as.factor(augment_value$delay)
augment_value$.predicted_label <- as.factor(augment_value$.predicted_label)
augment_value %>%
  accuracy(estimate = .predicted_label, truth = delay)
```

Make a prediction based on test data and get model performance.

```{r}
### Performance on test set
ml_predict(tree_model,test_data) %>%
  ml_metrics_multiclass()
```